\documentclass[11pt]{article}
\input{preamble}
\def\Poi{{\sf Poi}}

\begin{document}
\def\Exp{\mathbf{Exp}}
\def\Var{\mathbf{Var}}
\title{E0234: Assignment 9}
\author{}
\date{Due: Wednesday, March 30, 2016.}
\maketitle
It is highly recommended you do not google for the answers to the questions below. You can discuss with your friends, but then mention that in your submission.
The writing should solely be your own.

\begin{enumerate}
\item 
Show a strongly connected {\em directed} graph for which there exist vertices $u$ and $v$ such that the expected time to first reach $v$ starting from $u$ by a random walk is exponential in the number of vertices.

\item
Consider a graph $G$ known to be $3$-colorable. We want to $2$-color the graph such that no triangle is monochromatic (all three vertices have the same color).

Consider the following algorithm. First, color $G$ arbitrarily with red and blue. Then, as long as there is monochromatic triangle, pick a random vertex in that triangle and color it with the opposite color.

Show that the algorithm terminates in polynomial time. To do this, fix a legal $3$-coloring $\mathcal{C}$, and let $dis$ be the number of vertices in which the current $2$-coloring differs from $\mathcal{C}$. Analyze how $dis$ changes as the algorithm proceeds, and argue that if the algorithm takes too long, $dis$ can become $0$.

\item
Let $G_{n,N}$ be the uniform distribution on $n$-vertex graphs with exactly $N$ edges. Suppose $N = cn$ for some constant $c > 0$. Let $X$ be the expected number of isolated vertices (i.e.,~vertices of degree $0$) for a random graph from $G_{n,N}$. 
\begin{enumerate}
\item
Determine $\EX[X]$.
\item
Using Azuma, show that $\Pr[|X - \EX[X]| \geq 2 \lambda \sqrt{cn}] \leq 2e^{-\lambda^2/2}$. 
\end{enumerate}

\item
Consider a gambler playing a sequence of independent games, either winning one rupee with probability $1/2$ or losing one rupee with probability $1/2$. The gamber continues until either losing total of $\ell_1$ or winning total of $\ell_2$. Let $X_i$ be $+1$ if he wins in the $i$'th game and $-1$ otherwise. Let $Z_t = (\sum_{i=1}^t X_i)^2 - t$. 
\begin{enumerate}
\item
Show that $Z_1, Z_2, \dots$ form a martingale with respect to $X_1, X_2, \dots$.
\item
Let $T$ be the stopping time when the gambler finishes playing. What is $\EX[Z_T]$?
\item
Calculate $\EX[T]$. ({\em Hint}: Use what we showed in class about the probabilities of the player winner and losing at time $T$.)
 \end{enumerate}
  

\end{enumerate}

\end{document}