\documentclass[11pt]{article}
\input{preamble}
\usepackage{times}
\usepackage{graphicx}
\usepackage{algorithm,algorithmic}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{fullpage, prettyref}
\usepackage{pstricks,pst-node}
\usepackage{boxedminipage}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{ifthen}

\def\eps{\varepsilon}
\def\bar{\overline}
\def\floor#1{\lfloor {#1} \rfloor}
\def\ceil#1{\lceil {#1} \rceil}
\def\script#1{\mathcal{#1}}

\def\opt{{\tt opt}}
\def\alg{{\tt alg}}

\def\Pr{\mathbf{Pr}}
\def\Exp{\mathbf{Exp}}
\def\Var{\mathbf{Var}}
 
\setlength{\oddsidemargin}{0pt}
\setlength{\evensidemargin}{0pt}
\setlength{\textwidth}{6.5in}
\setlength{\topmargin}{0in}
\setlength{\textheight}{8.5in}
\setlength{\algobox}{6.5in}
\newcommand{\supp}{{\tt supp}}
\newcommand{\rank}{{\tt rank}}
\def\Pr{\mathbf{Pr}}
\def\Exp{\mathbf{Exp}}
\def\Var{\mathbf{Var}}

\begin{document}

\title{E0234 Randomized Algorithms}
\author{\bf Midterm: Solution Sketch}
\date{2nd March, 2016. 1:30pm to 4:30pm.}
\maketitle
\thispagestyle{empty}
\def\poly{{\tt poly}}
%You are allowed one A4 sheet of paper in your own handwriting which needs to be submitted with your answer sheet. 
\begin{center}
{\small 
Good luck!
}


\end{center}
%\hline

\begin{enumerate}
	

\item 
\begin{enumerate}
\item[(a)] Prove: for any two random variables $X,Y$, $\Exp[X+Y] = \Exp[X] + \Exp[Y]$.

Use basic definitions.

\item[(b)]
Prove or disprove: for independent random variables $X,Y$, \mbox{$\Var[XY] = \Var[X]\cdot \Var[Y]$.}

False. Take $X$ and $Y$ to be independent uniform random $0/1$ variables. $\Var[XY] = 3/16$ while $\Var[X] = \Var[Y] = 1/4$

\end{enumerate}
\vspace{0.5ex}

\item 
Consider a set system where each set has exactly $10$ elements and every element is present in exactly $10$ sets. 
Can you colour the elements red or blue such that every set has elements of both colours?

Yes. Use symmetric Lov\'asz Local Lemma.
%
%\item A hypergraph is $k$-uniform if each hyperedge has exactly $k$ vertices, and is called
%$k$-regular if every vertex is in exactly $k$ different hyperedges. A hypergraph is $2$-colorable if every vertex can be coloured either red or blue such that there are no monochromatic hyperedges. Prove that for $k>10$, every $k$-uniform, $k$-regular hypergraph is $2$-colourable.
\vspace{0.5ex}

\item Consider the random graph $G_{n,p}$ when $p = cn^{-2/3}$, and let $X$ be the number of $4$-cliques in the graph. 
\begin{enumerate}
	\item  What is $\Exp[X]$?
	
	$p^6 {n \choose 4}= \frac{c^6}{24}-o(1)$ 
	\item  What is an upper bound on $\Var[X]$?
	
	$\Var[X] \leq \frac{n^4}{24} p^6 + \frac{n^6}{8} p^11 + \frac{n^5}{6}p^9 \leq \frac{c^6}{24} + O(n^{-1})$
	\item  What is $\lim_{n\to \infty} \Pr[X = 0]$? 
	
	Use Janson's inequality to get that the limit is $e^{-c^6/24}$. 
\end{enumerate}
\vspace{0.5ex}

%\item Prove that for any two random variables $X$ and $Y$ (not necessarily independent), \\ $\Exp[XY] \leq \sqrt{\Exp[X^2]}\sqrt{\Exp[Y^2]}$.
%Assume $X$ and $Y$ are discrete.
%\item 
%$X$ is a random variable with expectation $\mu > 0$ and standard deviation $\sigma$. 
%Prove $\Pr[X > 0] \geq \frac{\mu^2}{\sigma^2 + \mu^2}$. You are allowed to use the statement from the previous question without proof.
%\vspace{0.5ex}
%\item Define  $S_n := |X_1 + \cdots + X_n|$ where each $X_i$'s are independent random variables which are  $\pm 1$ with probability $1/2$.
%Prove $\Exp[S_n] = \Theta(\sqrt{n})$.
%\vspace{0.5ex}
\item
Suppose you can draw independent samples of a real random variable $X$ that has expectation $0$ and standard deviation $\sigma$. Explain how to use only $O(\log n)$ samples from this source to generate a random variable $Y$ with expectation $\mu$ such that $\Pr[|Y-\mu| > 2 \sigma] < 1/n$. 

Let $S$ be set of $O(\log n)$ samples from $X$, and let $S' = S \cup -S$. Expected value of the median of $S'$ is $0$, and the median deviates from $0$ by more than $2\sigma$ with probability $1/n$. Finally, shift by $\mu$ to get desired variable.
\vspace{0.5ex}
\item 
Consider the following algorithm for the independent set problem on an $n$-node graph. Sample a random permutation $\sigma$ of $\{1,2,\ldots,n\}$. Initialize $I$ to $\emptyset$.
For $i=1$ to $n$, place $\sigma(i)$ in $I$ if it doesn't have an edge to any vertex in $I$.  In expectation, how large an independent set do you pick?

Expected size is $\sum_{i=1}^n \frac{1}{d_i+1}$ where $d_i$ is the degree of $i$'th vertex. Follows from the fact that probability $\sigma(i)$ is less than $\min\{\sigma(j): (i,j) \in E(G)\}$ equals $\frac{1}{d_i+1}$. 

\vspace{0.5ex}




\item  In $d$-dimensions, there can be at most $d$ unit vectors which are orthogonal to each other.
Call a pair of unit vectors $\eps$-orthogonal if $|u^\top v|\leq \eps$. How large (in cardinality) a set of pairwise $\eps$-orthogonal unit vectors can you construct in $d$-dimensions?

Can be as large as $k=\exp(O(\eps^2 d))$. Take a random unit vector $v_b = [\frac{b_1}{\sqrt{d}}, \frac{b_2}{\sqrt{d}}, \dots, \frac{b_d}{\sqrt{d}}]$ where each $b_i$ is $+1$ or $-1$ with equal probability. Then, for any two random $b$ and $b'$, $\Exp[v_b \cdot v_{b'}] = 0$ and $\Pr[|v_b \cdot v_{b'}| > \eps] < \frac{1}{k^3}$ by the Chernoff bound. We can then do a union bound over ${k \choose 2}$ pairs $b$ and $b'$ to get the claimed result.

\end{enumerate}

\end{document}